---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
always_allow_html: true
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators. A link to help you create a private repository is available on the #collaborative-project Slack channel. 

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.


**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: *vancouver_trees*
2: *flow_sample*
3: *apt_buildings*
4: *building_permits*

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->
Below, I will begin to explore 4 datasets to see what the data looks like in order to choose which I would like to work with. 

```{r}
# Exploring the *vancouver_trees* dataset
glimpse(vancouver_trees)
dim(vancouver_trees)
class(vancouver_trees)
```

```{r}
#Exploring the *flow_sample* dataset
glimpse(flow_sample)
dim(flow_sample)
class(flow_sample)
```

```{r}
#Exploring the *apt_buildings* dataset
glimpse(apt_buildings)
dim(apt_buildings)
class(apt_buildings)
```

```{r}
#Exploring the *building_permits* data set 
glimpse(building_permits)
dim(building_permits)
class(building_permits)
```

<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below ---------------------------->
I have narrowed down my datasets to *apt_buildings* and *vancouver_trees*. Although *apt_buildings* has many columns, they are all easy to understand what the variables are. The few that I do not know, including "rsn" seem to not be important enough for answering my possible research questions and can be filtered out. As for *vancouver_trees*, I am interested in environmental data for my thesis research so I may end up having similar variables in my thesis analysis work. However, this data set involves some complicated variables that I wouldn't know how to work with, including latitude and longitude. Further, this data set seems to more be a recorded account of where each tree is located for residential purposes versus for environmental conservation work, which makes me slightly less interested in it. 
<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below ---------------------------->
For the *apt_buildings* I could answer if private versus TCHC/Social Housing allows for pets or has better amenity options
For the *vancouver_trees* I could answer if the species correlates with the height and diameter of the trees

Based on these two possible research questions, I will choose *vancouver_trees* as that is a more interesting to me!
<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->
*Making a new tibble*

I will first make a new tibble with a subset of the data I am interested in by selecting the columns I will want to use. I will then assign these selections to a new data set name. I have chosen to do this step as I do not find every column relevant to my possible interests. Selecting for only the column I'm interested in makes the data set easier to work with. 

```{r}
#Assigning and making the new tibble
van_trees2 <-
(vancouver_trees %>% 
   select(tree_id, genus_name, species_name, common_name, neighbourhood_name, height_range_id, diameter, date_planted, longitude, latitude))

#Show the new tibble
head(van_trees2)
```
This simplified our table from 20 columns to 10, which will make it easier to work with.

*Investigating Missing Values*

Now, I would like to know about the missing value information in my data set, specifically, which columns have the most NA values, as this could affect my analysis. Below, I use tidyr functions to find the missing values (na's) across all the columns and then plotted this information. 

```{r Missing Values per Variable}
#Collecting the missing values per variable
missing_values <- van_trees2 %>% summarise(across(everything(), ~sum(is.na(.))))
#Plotting the missing values per variable
missing_values %>% pivot_longer(cols=everything()) %>%
  ggplot(aes(x=name, y=value))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, hjust=1))+
  xlab("Variable")+
  ylab("# of missing values (NAs)")+
  ggtitle("Number of Missing Values per Variable in *van_trees2*")+
  geom_text(aes(label=value), nudge_y = 3000)
```

This plot shows us that the variable "date_planted", representing when the trees were planted, is disproportionately missing values, which means we may want to exclude this variable from the analysis, or at least be aware of this fact when doing the analysis. Further, it makes sense to me that the latitude and longitude variables were missing the same amount of values as those are linked variables.

*Exploring the Relationship Between 2 Variables* 

Next, I was curious what relationships I could start to see between the variables. I know that older trees have larger diameters than their younger counter parts. I wanted to see if this pattern exists in planted residential trees by comparing the date planted and the diameter on a line graph. 

```{r Date Planted vs Diameter}
# Looking for date_planted and diameter correlation
van_trees2 %>% ggplot(mapping=aes(x = date_planted, y = diameter), na.rm = FALSE) + 
  geom_line() + 
  ylab("Diameter (DBH in inches)") + 
  xlab("Date Planted") +
  ggtitle("Date Planted vs Diameter")
```
There is no clear pattern here. However, we can see some obvious outliers in the diameter data, which may be interesting to look at later on. Further, we can also see some small gaps in time without trees being planted, including around 2010. However, this could be due to the missing values in dates_planted. 

*Plotting the Distribution of a Numeric Variable*

Additionally, as I start to think about what I can do with this data, I wondered if certain neighborhoods have significantly more trees planted there than others. Although I don't have data readily available on the wealth of these neighborhoods, it could be interesting to see if there is a link between trees in the neighborhood and cost of living. Let's look at which neighborhoods have the most trees. I will do this by grouping by the neighborhood_name variable and summarizing by the number of trees. I will then create a bar graph to better visualize this. 

```{r Trees Planted per Neighbourhood}
#Number of trees per neighbourhood in Vancouver
Neighbourhood <- van_trees2 %>% group_by(neighbourhood_name) %>% summarise(trees =n())

#Creating a bar graph of trees per neighborhood
Neighbourhood %>% 
  ggplot(mapping=aes(x = neighbourhood_name, y = trees), na.rm = FALSE) + 
  geom_bar(stat='identity') + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Trees Planted per Neighbourhood") +
  xlab("Neighbourhood in Vancouver") +
  ylab("# of Trees Planted")
```
By plotting this distribution, I found that Renfrew-Collingwood, Hastings-Sunrise, and Kensington-Cedar Cottage have had the most amount of trees planted since 1990. These numbers per neighborhood may, however, fluctuate over time. Further, there could be reasons for these neighborhoods having more trees planted in them, including the size of the neighborhood, lack of trees already being there, etc.


<!----------------------------------------------------------------------------->

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->
1) How have the number of trees planted in each neighborhood changed over time?
2) Are certain types of trees planted in some neighborhoods versus others?
3) How has species planted changed over time? 
4) How does tree size vary across neighborhoods? 
<!----------------------------------------------------------------------------->

# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.
8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
I will now look through each of my proposed research questions and start to summarize and graph the related information. 

**1) How have the number of trees planted in each neighborhood changed over time?**
```{r}
#Creating a categorical variable with 3 or more groups from an existing numerical variable

#Making a vector of dates
dates <- van_trees2$date_planted
#Extracting the year from dates as a new vector
year <- as.numeric(format(dates,'%Y'))

#Adding this vector into *van_trees2*
van_trees2$year_planted <- year


```
Creating this new variable was very helpful for my research question on trees planted over time as I made knowing just the year the trees were each planted accessible. This way I can look at trees per year. 


```{r Trees Planted Per Year}
#Creating a graph out of summarized variables that has at least two geom layers.

#I first need to count how many rows there are per year as this tells us how many trees were planted in each year
trees_per_year <- van_trees2 %>% count(year_planted)

#I then can create a bar graph of trees planted per year
ggplot(trees_per_year, aes(x=year_planted, y =n)) + 
  geom_col() +
  geom_text(aes(label=n), nudge_y = 100, size=2) + 
  ylim(0,4000) +
  xlab("Year") +
  ylab("Number of Trees Planted") +
  ggtitle("Number of Trees Planted Per Year in Vancouver") 
 


```

Creating a clear bar chart of year vs number of trees planted helped me clealry visualzie which years had the most trees planted. I can see that the graph is almost in a bell curve shape. 


**2) Are certain types of trees planted in some neighborhoods versus others?**
```{r}
#Computing the number of observations for at least one of my variables. Specifically, how many different tree species have been planted in each neighborhood
species_per_neighbourhood <- 
  van_trees2 %>%
  group_by(neighbourhood_name) %>%
  summarize(species = n_distinct(common_name))

head(species_per_neighbourhood)

```
Above, I calculated the number of species (by common_name) in each neighbourhood in Vancouver. This is very useful for my research question of looking at which trees are planted in which neighborhoods.

```{r Tree Species per Neighborhood}
#Creating a graph out of summarized variables that has at least two geom layers.

ggplot(species_per_neighbourhood, aes(x=neighbourhood_name, y = species)) + 
 geom_col() +  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(aes(label= species), nudge_y = 10, size=3)  +
  xlab("Vancouver Neighbourhood") +
  ylab("# of different tree species") +
  ggtitle("The number of tree species planted in each Vancouver neighbourhood") 

```
This graphic helped me visualize the diversity of trees per Vancouver neighbourhood, and showed me that there isn't a huge difference between neighbourhoods. However, it is interesting to see that Downtown has the least. I would later like to look at what the most popular tree type in downtown is. 

**3) How has species planted changed over time? **
```{r}
#Computing the number of observations for at least one of my categorical variables. Specifically, per year, how many different species were planted. 
species_per_year <- 
  van_trees2 %>%
  group_by(year_planted) %>%
  summarize(species = n_distinct(common_name))

head(species_per_year)

```
This shows if some years accounted for biodiversity more than other years. While it does not tell us which were the most popular per year, this is a good start.

```{r Tree Species per Year}
#Creating a graph out of summarized variables that has at least two geom layers.

ggplot(species_per_year, aes(x=year_planted, y = species))+
  geom_col() +  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_text(aes(label= species), nudge_y = 5, size=2.5)  +
  ylim(0,170) +
  xlab("Year") +
  ylab("# of different tree species") +
  ggtitle("The number of tree species planted per year in Vancouver ") 

```
This graph helps us visualize which years planted the most diverse trees. It is interesting to see that the diversity in tree species planted increased steadily over time since the start of planting on record until the late 1990s. I am not seeing many other interesting patterns here. It's important to keep in mind, however, that this isn't answering the entirety of the proposed research question. 

**4) How does tree size vary across neighborhoods?**
```{r}
#Based on two categorical variables, I will calculate two summary statistics
#First I will look at a summary of diameter
mean(van_trees2$diameter)
range(van_trees2$diameter)

#Next I will look at a summary of height_range_id (a category to denote height)
mean(van_trees2$height_range_id)
range(van_trees2$height_range_id)

```
This shows me the kinds of numbers I will be working with for this research question but it doesn't get me very far in answering my question.

```{r Height Range ID vs Diameter per Neighborhood}
#Making a graph with customized alpha transparency.
ggplot(van_trees2, aes(x=height_range_id, y=diameter)) +
 geom_point(aes(color = neighbourhood_name),
            size = 1,
            alpha = 0.8) 
```
This graph was not very useful for my research question as it is unclear with too many data points stacked on top of each other. This is due to trees being grouped into 10 height_range_id categories in the original data set. Perhaps then, I will only look at diameter when looking at size. 

<!----------------------------------------------------------------------------->

### 1.2 (3 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below ---------------------------->
For my first research question, I answered how the number of trees planted has changed over time in Vancouver but did not yet account for each neighbourhood. Here, I was able to get pretty close to answering my research question.
For my second research question, I  answered how many different species have been planted in each neighbourhood, but have not yet looked at the specific types of trees that are most popular in each neighbourhood. This is an interesting question to me as it looks at biodiversity per area. 
For my third research question, I was looking at how many different species were planted each year as this could tell us biodiversity planning over time. Again, I did not look at which species were most popular per year yet, but I did get a sense of overall biodiversity in species per year. 
For my fourth research question, I only scratched the surface for looking at tree size differences throughout the different neighbourhoods in Vancouver. I could refine this question to only look at one size variable (height or diameter). 
<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
